
# Omitting code  

Author: Seth Marshall

## Default fields

Functions in `R` often have default *fields* that occur in a given order, which can be found by looking up a function. For example, running `?ggplot` will open a help tab with information on how `ggplot()` is defined.

```{r, eval = FALSE}
library(tidyverse)
?ggplot
```

The default fields for `ggplot()` are as follows:

`ggplot(data = NULL, mapping = aes(), ..., environment = parent.frame())`

`data` is the first field, followed by `mapping`, etc. Since this is the default order for the fields, once you're familiar enough with coding to understand where each argument goes and why, you can omit certain fields.

```{r, eval = FALSE}
# example data
## recording the number of salamanders caught in pitfall traps per night
pitfall_trap <- c(8, 20, 6, 15, 9, 12, 5, 21, 6, 18, 16, 17, 3, 15, 23, 9, 10, 7, 30, 11,
                  11, 13, 7, 6, 18, 25, 0, 22, 14, 10)
## pitfall surveys were done nightly for a month
date <- c(1:30)

## making a data frame
pitfall_trap_data <-  data.frame(pitfall_trap, date)

# plotting with explicitly defined fields
ggplot(data = pitfall_trap_data, mapping = aes(x = date, y = pitfall_trap)) +
  geom_point() +
  labs(title = "Explicitly Defined Fields")

# plotting with omitted defaults
ggplot(pitfall_trap_data, aes(date, pitfall_trap)) +
  geom_point() +
  labs(title = "Omitted Defaults")
```

The two plots are the same, but the code is shorter in the second example.

Omitting fields when working with functions can be a powerful tool, but should only be used when you're familiar with what each argument serves to not cause confusion.

Additionally, field order can be changed by explicitly calling arguments, but is not compatible with omitting fields:

```{r, eval = FALSE}
# plotting with custom field order
ggplot(mapping = aes(date, pitfall_trap), pitfall_trap_data) +
  geom_point() +
  labs(title = "Custom Field Order")

# plotting with custom field order + omitting fields (error)
ggplot(aes(date, pitfall_trap), pitfall_trap_data) +
  geom_point() +
  labs(title = "Custom Field Order with Omitting (error)")
```
