# Lab 2: Indices in R, R Markdown  (2025)

Goals

In this lab, we will do a small amount of R programming, learn how to
write reproducible analyses using R Markdown, and get started on skills
to read primary literature. This week, rather than just reading the lab 
pdf, we want you to open the R Markdown file "WFC70_lab2.Rmd" and do your coding
directly in the relevant places in the Rmd.

By the end of this lab, you should be able to:

-   use indices and data frame column names to select particular pieces
    of data,
-   edit an R Markdown file to integrate text and code and change the
    document style,
-   knit an R Markdown file.


------------------------------------------------------------------------

## Part 1: R vs. RStudio, .R vs. .Rmd files (5 minutes, instructor leading)

### R vs. RStudio

This short section has some clarifications about the programs we are
use. You are using the programming language R. That is the primary skill
you are developing in these labs. But the program we open to write our R
code is called RStudio. RStudio is known as an IDE (an *integrated
development environment*), which is a fancy way to say that it is a
useful program in which to write and run your R code. So, on a future
CV, say that you are proficient in R (and, soon enough, R Markdown). You
don't have to explicitly mention RStudio.

This raises another issue that throws people off at the start. ***Always
begin your programming sessions by opening RStudio, not just the R
program.*** The default R program is not a very user friendly
environment for coding and does not work well with R Markdown (.Rmd)
files.

### R scripts (.R files) vs. R Markdown files (.Rmd)

When people are performing substantial amounts of R programming, they
tend to write their code in scripts. These are files that end in ".R",
and they are intended to include nothing but code (ideally with some
comments as well).

However, R scripts do not provide an easy way to combine regular text
and code. Sure, you can (and should) add comments to your code to
explain what each line is doing. But comments are not an easily readable
way to include large amounts of text. R scripts also do not allow for
images, plots, or displays of the results of code.

That's where R Markdown comes in. With a ".Rmd" file, the default is
that you are writing text. However, you can also add code chunks with R
code, and have that code run to generate interesting plots or other
outputs.

## Part 2: Indices and column names (5 minutes, instructor will lead)

Let's pull up last week's data very quickly.

```{r}
idaho_wolves <- c(1545, 1556, 1543, 1337)
idaho_years <- c(2019, 2020, 2021, 2022)

idaho <- data.frame(population = idaho_wolves, year = idaho_years)
```

It is often the case that you want to select only a particular column in
a data frame, or a particular value in a vector. Let's see first, then
discuss.

```{r}
idaho_wolves[4]

idaho$year
```

You should note that `idaho_wolves[4]` outputs the 4th value in that
vector (1337). And that `idaho$year` shows you the entire `year` column
in the data frame `idaho`.

Let's look at some more complex code.

```{r}
idaho_wolves[1:3]
idaho_wolves[c(1,3)]

idaho$year[2]
idaho[1,2]
```

A few things to unpack. First, we used a very odd function (the `:`
operator), which is a quick way for R to create a vector that is a
sequence of numbers. The code tells us to pull the 1st through 3rd entry in 
our vector. This function will often be useful. Try running `7:20`, then try 
running `10:5`. That's right, it even works backwards.

Next, if we want to specify several entries, we can't just write 1,3. We
need to make the relevant indices (1 and 3) into a vector as `c(1,3)`, with 
commas separating the values. Then we use that inside our square brackets to 
find the 1st and 3rd entry.

Because `idaho$year` is just a vector (from that column's values), we
can pick particular entries in the same way. However, if we have a
dataframe, there are both rows and columns. In that case, you would
entry two values/vectors: one for the row(s) and one for the column(s).
So `idaho[1,2]` gives you the value from the 1st row and second column.
We won't use this approach much -- there are more sophisticated tools
for selecting parts of our data for tidying or plotting.

## Part 3: Index work on your own (10 minutes, on your own)

Here are some new data.

```{r}
#creating a data frame for montana populations through time
mt_wolves <- c(1119, 1107, 1147, 1153, 1181, 1143, 1087)
mt_years <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022)
montana <- data.frame(population = mt_wolves,
                      year = mt_years,
                      state = c("montana"))

```

We did something cute, and added a third column that just says "montana"
in literally every row. (Nice to keep the state associated with the data
in the other columns.)

Your tasks:

1.  Write code to select the 2nd through 4th values in the `mt_wolves`
    vector.

2.  Write code to select the 1st, 2nd, and 6th entry in `mt_years`.

3.  Select just the `year` column in the `montana` data frame.

4.  Create a new `idaho` data frame that includes a population column, a
    year column, and a state column that just says "idaho" in every
    entry.

```{r}
# all code goes in here



```

## Part 4: Messing around with R Markdown (15 minutes, on your own)

In RStudio, if you click on the Help tab at the top, you will see a
selection labeled "Cheat Sheets". Select "R Markdown Cheat Sheet", which
will download and open a pdf. This cheat sheet is also posted on Canvas
in Files. I find it a little overwhelming, but here you really just need
to focus on the right side of the first page "Write with Markdown".

This section has instructions about how to format text, create headers,
make lists, etc. Let's try things out.

1.  **Knit this document to an html file.** To do this, click on the
    down arrow/triangle just to the right of the "Knit" button that is
    on the navigation bar below the name of the file. A dropdown menu
    will appear -- click "Knit to HTML".

This should quickly create an html file (like the kinds that represent
web pages) that you can view. It should include the text, code, and plots. 
Take a minute to see how different parts of the Rmd produced different kind of 
formatting.

Just above your Rmd file, you should also see a little button that says
"Visual". If you select that, you can switch to a visual edit which
allows for easy styling of an Rmd.

2.  **Create some headers using the \# symbol.** Which one is the
    biggest? The smallest? You can create headers of six different
    sizes, from \# to ######.

#### Put something here.

##### Put something else here.

###### And some more.

3.  **Style your text.**

Italicize this sentence.

Bold this sentence.

4.  **Create an unordered list** (like bullet points in a powerpoint) of
    your five favorite foods. For one of the foods, add at least two
    sub-items (perhaps those could be specific variations of that food)

5.  At the end of the previous section, we manipulated some Montana wolf
    data. **Use these numbers to create a table with two columns**
    (first header "year", second header "pop. size").
    
6.  Adjust the formatting of the next line so that it's not a header.

### Here I am writing a regular sentence, it should not be big and weird!


## Wrapping up (5 minutes)

Knit your file to HTML, and save the original Rmd, then add them to your
WFC 70 folder on Google Drive.

------------------------------------------------------------------------

------------------------------------------------------------------------

## Extra material for interest

-   Browse [this article on R
    Markdown](https://rmarkdown.rstudio.com/articles_intro.html). Try
    creating a new .Rmd file and exploring.

-   Try adding the option "echo=FALSE" to the top of an R code chunk
    (see the examples in the bottom left on the first page of the R
    Markdown cheat sheet). How does the HTML file look different after
    you knit the .Rmd file again? What about using "eval=FALSE"?

