
# Lab 3: Visualizing data in R (Part 1) (2025)


In this lab, we will get into one of the really fun parts of R programming --
***making figures!***

By the end of this lab, you should be able to:

-   load an R package using the `library()` function,
-   install R packages as needed using `install.packages()`,
-   use the `ggplot2` package within the `tidyverse` to write simple code to 
    plot data,
-   create boxplots, histograms, and scatterplots to compare the distributions
    of different subgroups of data.

### Resources

This lab is loosely based on chapter 1 in the online textbook "R for Data
Science" (second edition), which you can find online here:
<https://r4ds.hadley.nz/data-visualize>


--------------------------------------------------------------------------------

## Part 1: Packages in R (covered in lecture, so only here for reference)

Many functions and data in R are not instantly available. Instead, they are
available online as "packages". Packages extend the usefulness of R with new
functions and data (and documentation/metadata), and they are hosted online in a
standard way for easy installation and loading.

### Installing an R package

*Installing an R package* is easy to do within RStudio. In the computer labs we
have already installed most R packages that you will need for any coding in WFC
70. However, when you are programming on your personal computer you will need to
install any packages that need to be loaded for a particular lab or HW. ***You
only need to install an R package once.***

The code chunk below is only necessary if you are working on this lab at home,
because these packages are already installed in the computer lab.

```{r}
# the line below is commented so it won't run.
# if you need to install the packages, delete the hashtag then run 
# the line below, but then turn it into a comment again
# before knitting

# install.packages(c("tidyverse", "palmerpenguins"))


```

***Delete or comment out any install.packages() commands before you knit.
You cannot knit an Rmd that still has active install.packages() code.***

### Loading an R package

***On the other hand, you must have code to load any relevant R packages
inside of your Rmd if you go on to use functions/data from those packages
in your Rmd.***

*Loading an R package* tells R/RStudio that you plan to use that package during
your current sessions. ***You need to load an R package in every session that
you use it.*** After it is loaded, it will stay loaded until you close
R/RStudio.

```{r}
# loading the tidyverse package
library(tidyverse)
```

If you run this code and get the error message that there is no package called
'tidyverse', that means that the package is not installed, and you'll need to
use the earlier code to install it.

Assuming everything worked, in your output you'll notice a somewhat odd message.
It says "Attaching core tidyverse packages" and then lists several cryptic
names. Unlike most packages, `tidyverse` is a suite of multiple R packages, but
you can load them all in one line to save time, because they are commonly used
together.

Today we are primarily using functions from the `ggplot2` package, which was
loaded when we used the `library(tidyverse)` command in the code chunk above.

### Part 1b (5 minutes, instructor leading)

We also need to load the package with the penguin data.

```{r}
#loading the palmerpenguins package, which contains two data frames
library(palmerpenguins)
```

There is one odd thing about data sets that are included in R packages. You will
notice that nothing new has shown up in the global environment. However, the
data frames are actually loaded (but hidden).

```{r}
penguins
penguins_raw
```

I prefer to formally load the data into my environment, which you can do with
the `data()` function.

```{r}
# the palmerpenguins package called their data "penguins"
# and the command will load both data frames
data(penguins)
```

Now we can see the data frames in our environment. Let's take a closer look.

```{r}
# the glimpse() function lets you see the variable (columns) and
# a few observations for each variable (rows)
glimpse(penguins)
glimpse(penguins_raw)
```

A few questions to orient ourselves to these data.

1.  Which data frame has more variables, or are they the same?

2.  Which data frame has more observations, or are they the same?

3.  What is different about the column/variable names in `penguins` as compared
    to `penguins_raw`?


## Part 2: Using ggplot() for plotting in R (15 minutes, self-led)

The core function in the `ggplot2` package is called `ggplot()`. But making
plots using `ggplot()` is like making lasagna. You have to add layers until
you've produced the perfect result.

Recall from lecture that we need the base canvas with the data (and aesthetics),
as well as at least one additional *geometry* layer. We will see lots of types 
of `geom` in this course. We'll briefly begin with a critical type of plot, 
the scatterplot. That shows two-dimensional points to illustrate the 
relationship between the x- and y-axis. It has a geometry called `geom_point()`.

```{r}
ggplot(data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

Hooray, we have made our first ggplot! However, today we want to practice
making histograms and boxplots. So we'll need some slightly different aesthetics
and geometries. We only started with scatterplots because they are so common,
and they have a very clearly defined x- and y-axis.

***Before we move on, there is one crucial thing to note: each time you add a
new layer (e.g. a `geom`, some labels for nicer x- and y-axis titles, etc.) you
need to have a plus symbol, +, at the end of the previous line of code.
Otherwise R does not know to keep looking for additional layers before making
the plot.***

Let's try out a histogram. First, we might simply want to visualize the
distribution of all of our measurements of bill length.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm)) +
  geom_histogram()
```

Because histograms just count values for one variable, there is only
one aesthetic (`x=bill_length_mm`). The geometry was for a histogram is
`geom_histogram()`. There are a few things we might want to consider.

-   This has a lot of separate bins. We might say it is a very *fine-grained*
    histogram. We can adjust that with *arguments* inside of `geom_histogram()`.
-   This is summarizing data from multiple species at once. That is not very
    meaningful. We can separate these out in several different ways.

To address the first issue, you may have noticed that geometries are also
functions, so their output can be modified using arguments just like most other
functions in R. By default this function creates 30 bins (thirty separate
x-values/bars) for a histogram. Let's specify a smaller number.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm)) +
  geom_histogram(bins = 20)
```

Alternatively, we could specify a `binwidth`, setting how wide each bin should
be. Let's try using a `binwidth` of 2 (the units for this column are millimeters.)

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 2)
```

Now for the fun part. We can quickly add informative complexity. Let's keep 
track of all three species in our histogram. We can add the `fill` aesthetic. 
`fill` and the very similar `color` aesthetics allow you to color code different
data points based on another variable/column in your data. Let's try `color` 
first and see why it is NOT ideal here.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm, color = species)) +
  geom_histogram(binwidth = 2)
```

Hmmm, this is not right. We want the whole bars colored. For some geometries,
`color` only changes the borders, not the full shapes. But color will often be
useful, for example to change the color of points when using `geom_point()`. To
fill in a whole shape with color, you use `fill` instead.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm, fill = species)) +
  geom_histogram(binwidth = 2)
```

This is still not quite right. It is showing all the data, now color-coded, but
it is stacking the numbers rather than showing three separate histograms with
some overlap in the x-values. That is because `geom_histogram()` defaults to
stacking data, even if there is a `fill` aesthetic. You can fix that by changing
another argument called `position`.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm, fill = species)) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.7)
```

We made two changes. We made it so each species has a separate
histogram starting from a count of 0. And we made the bars partially transparent
with the `alpha` argument. (0 is totally transparent, 1 is totally solid
colors.) It's a little wonky, but we can pretty much read each histogram now.

Instead, let's switch to box-and-whisker plots. These use a different geometry,
`geom_boxplot`, and they will have a continuous variable for one axis, and
usually also a discrete variable (e.g., species) for the other axis. Let's
stick with bill length and species and make three vertical boxplots, one for
each species.

```{r}
ggplot(data = penguins,
  mapping = aes(x = species, y = bill_length_mm)) +
  geom_boxplot()
```

This is easier to look at. We can see quite clearly that Adelie Penguins have
the lowest bill length, while the Chinstrap and Gentoo Penguin samples have
subtantial overlap in bill length.

## Part 3: Exploring (15 minutes, self-led)

Use the code chunks below to try out some plotting with a new variable.

1.  Create a histogram like the fancy, partially transparent one we made above,
    but for the variable `bill_depth_mm` instead of bill length, still using the
    `fill` aesthetic with the `species` column. Hint: copy and paste that 
    code into the chunk below, then switch the variable name(s) as needed.

```{r}

# write your code in this chunk

```


2.  Now create a boxplot with `species` as the x-axis and `bill_depth_mm` as the
    y-axis.

```{r}

# write your code in this chunk

```

3.  Compare with your results for bill length. Bill depth refers to the
    thickness (top to bottom) of a bill, while bill length is the distance from
    the base of the bill to the tip. Which species seems to have an overall
    short, thick bill? Which species seems to have an overall long, skinny bill?
    Double-check by googling the species to see some photos.

*Remove this and add your text notes here.*

4.  Try to fix these snippets of broken code. You will note that we have added
    an option to these code chunks that says "eval = FALSE". That is how we say
    that RStudio should not try to actually run (evaluate) that code when
    knitting the Rmd file into HTML. Because the code is broken, it you tried to
    knit when "eval = TRUE" (i.e. when RStudio will try to run that code), then
    it will return and error and your code won't knit. Feel free to remove that
    option or switch it to "eval = TRUE" after you have fixed the code for each
    chunk. ***Note: if you are ever trying to knit a HW and your code is
    returning errors, you can always set those chunks to "eval = FALSE" if you
    want to still be able to submit your knit file.***

Chunk 1 -- this plot won't run.

```{r}
# uncomment the code below and try to fix it

#ggplot(data = penguins,
#  x = species,
#  y = body_mass_g) +
#  geom_boxplot()
```

Chunk 2 -- a blank plot won't do.

```{r}
ggplot(data = penguins,
  mapping = aes(x = species, y = bill_depth_mm))
  geom_violin()
```

5.  **(Time permitting.)** The code below creates a scatterplot with bill length
    and bill depth, while also color coding the points. Review the code, then
    try to do the same thing for a boxplot of bill length, color-coding each box
    for the species it represents.

```{r}
ggplot(data = penguins,
  mapping = aes(x = bill_length_mm, 
                y = bill_depth_mm,
                color = species)) +
  geom_point()

```

## Wrapping up (5 minutes)

Knit your file to HTML, and save the original Rmd, then add them to your WFC 98
folder on Google Drive.

--------------------------------------------------------------------------------

--------------------------------------------------------------------------------

## Extra material

-   Look over the intro chapter on data visualization in the R for Data Science
    textbook <https://r4ds.hadley.nz/data-visualize>. Here are a few things to
    try out.
    -   Add a `geom_smooth()` layer on top of `geom_point()` for one of the
        scatterplots we made. What does it do?
    -   Check out `geom_density()` instead of `geom_histogram()` in one of our
        histogram-making code chunks above. What does it do? Try adding the
        argument `adjust = 1/2` or `adjust = 2` and see how it changes the
        shape. (You will need to remove the `bin` or `binwidth` options, because
        those are only relevant for `geom_histogram()`).
    -   Try out the exercises in section 2.4.3.
-   Try to make the same scatterplot we made in our very first plot, but using
    `penguins_raw` instead of penguins. Can R handle column names that have
    spaces in them? (Hint: using the left quote symbol, \`, on either side of
    the name will allow R to understand that you are trying to refer to a column
    name.)
-   Check out this code and plot below. It is a prettier version of our boxplot for
    culmen length. Google some of the additional layers I added to see how to 
    create and modify them yourself.
    You may want to explore themes, explained more on [this website](https://ggplot2.tidyverse.org/reference/ggtheme.html). Rob often finds
    himself using `theme_bw()` and `theme_classic()`.

```{r}
ggplot(data = penguins,
  mapping = aes(x = species, y = bill_length_mm)) +
  geom_boxplot() +
  geom_jitter(width = .1) +
  theme_classic() +
  labs(x = "", y = "culmen length (mm)")
```
