
# Working with Spatial Data

Author: Mingyuan Li

```{r, eval = FALSE}
# We are using multiple new packages, install and load them if you haven't done it so:
# install.packages(c("move", "move2", "sf", "raster", "terra", "sp", "adehabitatHR", "amt", "mapview"))
library(move)
library(move2)
library(sf)
library(sp)
library(adehabitatHR)
library(amt)
library(mapview)
library(raster)
library(terra)
library(tidyverse)
```

Before we start, let's talk about GIS!
GIS stands for geographic information system, it help us analyze and visualize data that
are spatial in nature.

What is the different between spatial data and other data types? Ultimately, not much!
Just like other data types, spatial locations can be depicted on x and y axes.
But if we were to include things like elevation, then we can use 3-dimensional
data analysis and visualization techniques.

Two things that sets spatial data apart from other data types:
 1. In non-spatial data, we care about statistical relationships(How variable X influence variable Y)
    In spatial data, we care more about location-based relationships(How points/locations are related in space)
 2. Spatial data are often organized in multiple layers, so for example we can have a location with
    layers elevation, temperature, vegetation coverage... and they stack together.

Today, we will start by directly download a dataset from Movebank:

In R, by using "move" and "move2"package, we can directly download data sets from MoveBank.
But first, you need a Movebank account to login. So open and past this link into your browser:
https://www.movebank.org/cms/movebank-main
Use "Login/Register" at top right to login.

```{r, eval = FALSE}
# Now you can login directly through RStudio:
loginStored <- movebankLogin(username="Your_username", password="Your_password")
```


Next, we need to find a study that we want to explore in R,
For our purpose of the lab, we are using cheetah data for demonstration
To do that, go back to Movebank page, in "Data" section, go to "Explore map", and on the left you should see a search bar, check the box "Only studies where I can see data" and enter "Cheetah Pilanesberg National Park"
Next, click the box with "i" that's next to the study name and select "Open in studies page"
Here is all the information you need to load the dataset!

Before moving to R, you need to check the terms and conditions, in Movebank study page, click "Download data", and agree to the
terms and conditions. Once agreed, you are all set to continue!

```{r, eval = FALSE}
# Now in the study page, find the "Movebank ID" and enter it in the code below next to "study = "
cheetah_move <- getMovebankData(study = , login = loginStored)
### The study number should be 1260886163

#It's always a good idea to check the data structure before we began
view(cheetah_dat)
str(cheetah_move)
# What is the class of the data?

# Let's transform the data from a MoveStack object into a dataframe
cheetah_dat <- as.data.frame(cheetah_move)

head(cheetah_move)
# What kind of information do we have?

# Your turn! Now use select() or dplyr::select() function to create a new data frame called "cheetah_clean" that includes columns of longitude, latitude, tag local identifier, timestamps, and sex.
# Remember you can always use view()/head() function to obtain the column names

### it should be something like this:
# cheetah_clean <- cheetah_dat %>%
#   dplyr::select(
#     tag_local_identifier,
#     location_lat.1,
#     location_long.1,
#     timestamps,
#     sex)


```

## Understanding coordinate reference systems(CRS)

In spatial R, we use different coordinate reference systems(CRS) to define where and how the data is on Earth.
The two most common CRS In R is WGS84 and UTM system, but what is the difference?
Well, WGS84 (World Geodetic System 1984) system are stored as latitude/longitude (lat/lon) in units of degrees, while UTM projects everything on a flat surface in units of meters (x/y) and divided Earth into 60 grids we refereed as UTM Zones.
This difference make WGS84 work well with global datasets/GPS on a larger scale, but it will create distortion near polar regions and on smaller scales.
Remember unlike WGS84, UTM data are stored in meters and uses a grid system, this minimized distortion and make it great for precise local analysis.
Here is a table for better comparison:
| Feature        | WGS84                     | UTM Zones                       |
|----------------|---------------------------|---------------------------------|
| Units          | Degrees (°)               | Meters (m)                      |
| Best For       | Global                    | Local analysis                  |
| Distortion     | High near poles           | Minimal within zone             |
| Data Structure | Lat/Lon                   | X/Y grid                        |
Also, be sure to always check your CRS in R with st_crs() or crs()!!! Mixing systems can lead to errors!!

Now going back to our cheetah example, to visualize our data on a map, we need to tell R which CRS we want to use.
To do so, we can use "sf" or "sp" package, both package create spatial layers from vector data.
However,"sf" is relatively new, much newer than "sp". Which "sp" is the older way of dealing with
spatial vector data in R. That means, some older packages only work with "sp" data, not
"sf" data. This includes home range package(adehabitatHR) we are going to use later today, but ggplot only works with "sf" data! So we need to switch back and forth.


Let's try it with "sf" package first!
```{r, eval = FALSE}
# We can convert our cleaned data to an "sf" object using st_as_sf()
# For coordinates, be sure to put longitude first!
cheetah_sf <- st_as_sf(cheetah_clean, 
                     coords = c("location_long.1", "location_lat.1"), 
                     crs = 4326)
# Note the number "4326" is the EPSG code for WGS84, which tells R how to interpret the coordinates.

# Let's check the class of the new dataframe, it should say "sf" "data.frame"
class(cheetah_sf)

# Now we can use the sf object we created and visualize with ggplot!
ggplot() +
  geom_sf(data = cheetah_sf, color = "orange", alpha = 0.5, size = 1.5) +
  labs(title = "Cheetah Locations (WGS84)") +
  theme_bw()

# sf object also supports fancy interactive maps liken mapview package, try zoom in/out and change map layers!
mapview(cheetah_sf)

```

"sp" package and home range

```{r, eval = FALSE}
# Let's first create a copy from our clean data
cheetah_sp <- cheetah_clean
```

For home range estimation (which home range is the area that's regularly used by animal), requires units in meter, so we will use UTM system for CRS.

Unlike with the package "sf" that we just used, in "sp" package, we need to assign the 
coordinates using "coordinates()" and the crs using "proj4string()".
Unfortunately, "sp" doesn't work nearly as well with the tidyverse as "sf", so we have to run the code one at a time instead of writing a pipe.

```{r, eval = FALSE}
# First, set the coordinates, and Longitude always go before latitude!
coordinates(cheetah_sp) <- ~ location_long.1 + location_lat.1


# Remember, for UTMs, we need to specify the zone. From the project summery we know that Pilanesberg National Park is in Zone 35 S (South), and now we can manually assigned the coordinates using proj4string().
proj4string(cheetah_sp) <- CRS("+proj=longlat +datum=WGS84")

# Note this is still in WGS84, so we need to transform to UTM Zone 35S to avoid errors
cheetah_sp_utm <- spTransform(cheetah_sp, 
                            CRS("+proj=utm +zone=35 +south +datum=WGS84 +units=m +no_defs"))

# Check the class of the new dataframe
class(cheetah_sp_utm)
# Now the data should be in SpatialPointsDataFrame
```

Let's now find the home range for our cheetah using minim!m convex polygon (MCP)!
MCP basically the smallest convex polygon that contains the percentage of area used by animals, it's a tool that everyone like to use in ecology, and it's fun to visulize!
(You don't really have to understand this, but it help picturing the application of spatial R!)

```{r, eval = FALSE}
# Here we can use the mcp.area() function to calculate the home range from 0 to 100%
cheetah_mcp_areas_sp <- mcp.area(cheetah_sp_utm[1], unin = "m", unout = "km2")

# Which then we can take a quick look and plot them out
plot(cheetah_mcp_areas_sp)
```

Want to visulize the the MCPs with ggplot?
Remember spatial points(sp) cannot be directly plotted using ggplot, so we need to convert them to sf objects before plotting

```{r, eval = FALSE}
# We can also visualize the 95% and 50% mcp by calculate the home range individually
cheetah_mcp_95_sp <- mcp(cheetah_sp_utm[1], percent = 95)
cheetah_mcp_50_sp <- mcp(cheetah_sp_utm[1], percent = 50)

# Yes you can convert the data to sf objects before plotting, but here we will convert them inside the ggplot to save space
ggplot() +
  geom_sf(data = st_as_sf(cheetah_mcp_95_sp), fill = "purple", alpha = 0.5) +
  geom_sf(data = st_as_sf(cheetah_mcp_50_sp), fill = "orange", alpha = 0.5) +
  geom_sf(data = st_as_sf(cheetah_sp_utm), color = "black", size = 1) +
  labs(title = "95% and 50% MCP Home Ranges for Cheetah CF01") +
  theme_bw()

# Also try using the fancy mapview to take a closer look
mapview(st_as_sf(cheetah_mcp_95_sp), col.regions = "purple", alpha.regions = 0.5, layer.name = "95% MCP") + 
  mapview(st_as_sf(cheetah_mcp_50_sp), col.regions = "orange", alpha.regions = 0.5, layer.name = "50% MCP") +
  mapview(st_as_sf(cheetah_sp_utm), col.regions = "black", cex = 3, layer.name = "GPS Locations")

```

Your turn!
Now we are going to work on another data set from a puma project in Utah
Try using the code we already used!


```{r, eval = FALSE}

# First, download the puma data from Movebank, I already had the study number for you this time
puma_move <- getMovebankData(study = 1720694224, login = loginStored)

#1. Now convert the our puma_move object to a dataframe called "puma_dat" using as.data.frame() function

## puma_dat <- as.data.frame(puma_move)


#2. Use dplyr::select() function to create a new data frame called "puma_clean" that includes columns of longitude, latitude, tag local identifier, timestamps, also use dplyr::filter() to filter out only puma in the taxonomy species.
# Hint: you can use view()/head() to see the column names


# puma_clean <- puma_dat %>%
#   dplyr::filter(taxon_canonical_name == "Puma concolor") %>%
#   dplyr::select(
#     local_identifier,
#     location_lat,
#     location_long,
#     timestamps)


#3. Next, use st_as_sf() to convert our cleaned data to sfobject called "puma_sf", and set crs to 4326
# puma_sf <- st_as_sf(puma_clean, 
#                    coords = c("location_long", "location_lat"), 
#                    crs = 4326)

#4. Now creat a ggplot to visualize the data initially, for sf data, you should use geom_sf(), you can also try mapview() for a better look
# ggplot() +
#   geom_sf(data = puma_sf) +
#   theme_bw()


#5. Let's convert to sp object for home range analysis
# a) Create a copy of puma_clean called "puma_sp"
# b) Use coordinates() to set spatial coordinates
# c) Use proj4string() to set WGS84 CRS
# puma_sp <- puma_clean
# coordinates(puma_sp) <- ~ location_long + location_lat
# proj4string(puma_sp) <- CRS("+proj=longlat +datum=WGS84")

# Next, we will transform to UTMs and create MCP polygons, since Utah is in UTM Zone 12N, we will use that instead.
# Here I had the code for you, so no worries.
puma_sp_utm <- spTransform(puma_sp, 
                         CRS("+proj=utm +zone=12 +datum=WGS84 +units=m +no_defs"))
puma_mcp_95_sp <- mcp(puma_sp_utm[1], percent = 95)
puma_mcp_50_sp <- mcp(puma_sp_utm[1], percent = 50)

#6. Now, make a ggplot and/or mapview to visulize the MCPs, remember you need to convert the data to sf objects first!
# ggplot() +
#   geom_sf(data = st_as_sf(puma_mcp_95_sp), fill = "blue", alpha = 0.5) +
#   geom_sf(data = st_as_sf(puma_mcp_50_sp), fill = "lightblue", alpha = 0.5) +
#   geom_sf(data = st_as_sf(puma_sp_utm), color = "red", size = 0.5, alpha = 0.5) +
#   theme_bw()

# mapview(st_as_sf(puma_mcp_95_sp), col.regions = "blue", layer.name = "95% MCP") + 
#   mapview(st_as_sf(puma_mcp_50_sp), col.regions = "lightblue", layer.name = "50% MCP") +
#   mapview(st_as_sf(puma_sp_utm), col.regions = "red",layer.name = "GPS Locations")
```




Credits for this lab:
1. University of California, Davis, course WFC 155 lab 5 and 8 
2. Intro to spatial R by Claudia A Engel: https://cengel.github.io/R-spatial/intro.html

