---
title: "Lab 5: data story work, wrangling warm-up"
author: "WFC 70, Fall 2025"
output: html_document
---


## Goals

In this lab, we will do a tiny big of wrangling, but mostly focus on getting 
things started for the Data Story. More wrangling to come in the next two weeks.

By the end of this lab, you should be able to:

-   use `filter()` to select of subset of a data frame,
-   add faceting to plots,
-   describe the potential wrangling process for a data set for your Data Story.

### Resources

We will work with a new data set today based on small mammal trapping to
assess biodiversity in different types of plots. These data are explored
in detail in the Data Carpentry course "Data Analysis and Visualisation in R for Ecologists", [https://datacarpentry.org/R-ecology-lesson/index.html](https://datacarpentry.org/R-ecology-lesson/index.html). 



## Part 0: Figure of the week (5 minutes, instructor leading)

--------------------------------------------------------------------------------

--------------------------------------------------------------------------------

## Part 1: New data! (5 minutes, instructor leading)

This is mostly review about the data introduced at the end of Monday's lecture.

### Loading data

```{r}
# loading a csv file from this URL. The header argument just tells R
# that the first row is not data, but is instead a header
# with column names
mammals <- read.csv("https://raw.githubusercontent.com/refurrow/WFC70_old/main/portal_data_joined.csv",
                    header = TRUE)

# loading tidyverse for our plotting and data wrangling needs
library(tidyverse)
```

### About these data
The data are part of the Portal Project Teaching Database [https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10](https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10).
They come from a long-term mammal and plant monitoring study in the Chihuahuan 
Desert near Portal, Arizona.

```{r}
# getting a sense of our variables
glimpse(mammals)
```

You'll notice that we have a lot of data. The columns tell us the day, month,
and year that each observation was made, as well as the plot id, the genus and
species of each observation, a `species_id` that is a shorthand version of 
the species name, the sex, weight, and hindfoot length, and whether the 
species is a rodent, bird, rabbit, or reptile. 

Finally, `plot_type` tells us whether it was a control plot (no manipulation),
a plot where they excluded kangaroo rats for an extended period (long-term), a
plot with short-term exclusion of kangaroo rats, a plot where all rodents
were excluded, or a plot where only one species was excluded
(*Dipodomys spectabilis*).

## Part 1: visualizing size relationships

Your first activity is to explore size relationships on a filtered subset
of these data, looking at just a few species. We will get you started with
some code to filter the data.

```{r}
# selecting only genus Dipodomys, and only for the control plots
# then removing any individuals who were not identified to species (the "sp.")
dipodomys <- filter(mammals, genus == "Dipodomys" & plot_type == "Control") |>
  filter(species != "sp.")
```

We will learn more about the pipe, `|>`, next week. It is basically saying 
"Then do this", and passing your data frame on to the next line. It allows you 
to not have to re-enter the name of the data frame at every step of tidying
data. Here, after we filtered to only genus *Dipodomys* and only control plots,
we then did a second filter to remove any individuals who were just labeled
as *Dipodomys sp.* rather than the name of a specific species. (When you see
*sp.* for a species name that means that the individual was left unidentified.)

We now have a simplified data frame for only the three kangaroo rat species. 
Here are two examples of visualizing the hindfoot and weight relationships for 
each species.

```{r}
# scatterplot with color
ggplot(data = dipodomys,
       mapping = aes(x = hindfoot_length, y = weight, color = species_id)) +
  geom_point()

# faceting by species
ggplot(data = dipodomys,
       mapping = aes(x = hindfoot_length, y = weight)) +
  geom_point() +
  facet_wrap(~species)

```

### Your task (10 minutes)

Starting from the `mammals` data frame, use `filter()` to make a new data
frame called `sigmodon` that includes only species in the genus Sigmodon. With
these new data, create a violin plot of hindfoot length, colored by sex, and
faceted by species. You should find the code above to be very helpful.

```{r}

# write your code and comments in this chunk

```


## Part 2: a wrangling plan

You proposed some potential data sets for your Data Story in your HW 3
submission. Use the rest of today's class to explore at least one of these
data sets more deeply.

In particular, focus today on what kind of data wrangling may need to be done
in order to get your data ready for making your Data Story plots. Here are
a few guiding questions.

* What are the columns of interest in my data? What type of data is each column
(e.g., numeric, categorical, text)?

* Is the data structured in a "long" format for use with ggplot(), where
each row is an observation?

* Do the data need cleaning? Are there rows that need to be filtered out?
Are there columns that have a mix of numeric and text data (which will make
R force the whole column to be character/text data)?

* Do the data need any enriching/transformation? Do I want to make a new 
column based on data in other columns? Do I need to combine two or more data
frames (e.g., combining weather data with mammal trapping data for each night)?

* Do I need to summarize the data in any way, e.g., by calculating means and
standard errors for certain ages or by sex?

* How can I validate the data to catch any errors with my coding, or errors
with the raw data?  (Make a plan of certain plots that you can use to 
double-check that the data generally make sense and show the expected patterns.)



***

Write your notes/answers here for future reference. You may end up using some 
of this text when writing your Data Story. 

***


## Wrapping up (5 minutes)

Knit your file to HTML, and save the original Rmd, then add them to your WFC 98
folder on Google Drive.


***

## Extra material

1. Work on HW 4.

2. Try out the exploratory data analysis activity below.


### Transforming the data (scary code behind the scenes)

Run the code below to create a new data frame that summarizes the diversity 
at each plot for each year.

```{r}
# creating a function to calculate Shannon's diversity index
shannon <- function(x)
{
  proportions <- table(x)/length(x)
  D <- -sum(proportions*log(proportions))
  return(D)
}

# calculating the diversity of rodents in two ways, after filtering data
mammals_diversity <- mammals |>
  filter(taxa == "Rodent" & genus != "Dipodomys") |>
  filter(genus != "Rodent" & species != "sp.") |>
  group_by(plot_id, year) |>
  summarize(richness = length(unique(species_id)),
            shannon_d = shannon(species_id)) |>
  mutate(plot_type = case_match(plot_id,
                                c(2,4,8,11,12,14,17,22) ~ "control",
                                c(3,15,19,21) ~ "krat long-term exclosure",
                                c(5,7,10,16,23,24) ~ "all rodent exclosure",
                                c(6,13,18,20) ~ "krat short-term exclosure",
                                c(1,9) ~ "spectab exclosure"))
```

We have two different measurements of diversity. The `richness` column
is just a count of exactly how many different rodent species were found 
in that plot in that year.  The `shannon_d` column is a measure of 
Shannon's diversity index. That value quantifies diversity in a way that
includes species richness but also species evenness (whether there were
many common species, or just a few common species and many rare species).  
Ecologists generally consider a community to be more diverse if there are 
many common species, rather than just a few common species and lots of rare 
species.

### Your task

Using the mammals diversity data, try to identify if there are differences 
in diversity based on plot type, if there have been changes in diversity 
through time, or both. Try plotting the data in many ways, with different
plot types and different combinations of which variables are shown using
x-values, y-values, colors, facets, etc. Remember that we have two different
measures of diversity. This kind of quick plotting in many ways is a key part 
of exploratory data analysis.

***Create an effective figure to illustrate a pattern that you have found.***

```{r}

# enter your code and comments here

```




