
# Lab 6: mark-recapture (2025)

## Goals

In this lab, we will use our wrangling skills to clean up a data frame
in order to estimate population sizes using mark-recapture. 

By the end of this lab, you should be able to:

-   calculate a population estimate from mark-recapture data,
-   use one of the `_join()` functions to combine different data frames that
have a shared "key" column, 
-   use `bind_rows()` to combine multiple data frames with 
identical columns.

### Resources

We will use the data discussed during the mark-recapture lectures. These are
data from a LTER (Long-term Ecological Research) site called the San Joaquin
Experimental Range.  
[https://data.neonscience.org/data-products/DP1.10072.001](https://data.neonscience.org/data-products/DP1.10072.001)


## Part 0: Figure of the week (5 minutes, instructor leading)

--------------------------------------------------------------------------------

--------------------------------------------------------------------------------

## Part 1: Reminder of lecture data (10 minutes, instructor leading)

This code should align with what we covered during lecture. Note, however,
that you don't even have to do the Schnabel calculation by hand. With the
right vectors for M, T, and R, that final line does the multiplying, adding,
and dividing, and should yield exactly what we found during lecture.

```{r}
# loading tidyverse for our plotting and data wrangling needs
library(tidyverse)

# loading data
sjer_jan17 <- read.csv("https://raw.githubusercontent.com/refurrow/WFC70_old/refs/heads/main/data/mam_pertrapnight_2017_01_basic.csv")

# wrangling for a simpler data set with just one species
pebo_jan17 <- sjer_jan17 |>  #start with the full data, THEN
  filter(scientificName == "Peromyscus boylii") |> #filter for P. boylii, THEN
  select(collectDate, tagID, recapture, fate, plotID, nlcdClass) # selecting only these columns

# creating a table from the date column and the recapture column
pebo_table <- table(pebo_jan17$collectDate,pebo_jan17$recapture)
pebo_table

# entering our mark-recapture data
pebo_M <- c(4, 14)
pebo_T <- c(12, 6)
pebo_R <- c(2, 4)

# adding up sum of M*T values, then dividing by sum of R values
sum(pebo_M*pebo_T)/sum(pebo_R)
```

You estimate 22 individuals in this population, using the Schnabel estimator.
Okay, in a moment we will set you loose to try things on your own, but
first let's do a quick `_join()`.

## Adding weather data (joining data frames)

Small mammal activity at night may depend on nighttime temperature or
precipitation, so it would be interesting to see whether our sample size
varies depending on the weather conditions.  (For example, if our mice
are inactive in very cold weather, we will expect to catch fewer mice in 
our traps on cold nights.)


```{r}

# weather for the site
# the csv was having formatting issues, so we switched here to loading
# a txt file using read.delim()
sjer_weather <- read.delim("https://raw.githubusercontent.com/refurrow/WFC70_old/refs/heads/main/sjer_weather.txt")
```

We have many fewer rows, because for each night we just have a single
suite of measurements (rather than separate rows for each individual
caught in our original data). `lowTemp` is the lowest temperature that night
in degrees Fahrenheit, and `precip` is the overall rainfall in the previous
24 hours, in inches.

To think about overall activity, let's not simply focus on *P. boylii*. Let's 
instead use all of our data. We'll want to remove all the rows with
empty traps and only include rows with captures. There are many ways
we could filter here. I noticed that the when a trap is empty the recapture 
column is always the empty string, "", while the column always has a value when 
an individual was caught, so we can filter by removing all the empty rows. The 
exclamation point, !, means NOT, so we will filter the rows that are NOT equal 
to "".

```{r}
# getting all catches (not just P. boylii)
catches_jan17 <- sjer_jan17 |>
  filter(recapture != "") # only rows that are NOT equal to the empty string, "".
```

Now, let's join. We don't care about the February or October weather, so let's 
prioritize the rows in our catch data. 

* ***Pause and think. What is our shared key? What kind of join should we use?***

```{r}
#joining catch data and weather data
catches_jan17_full <- left_join(catches_jan17,sjer_weather, by = "collectDate")

```

* ***Pause and check. Do the data look like what we expected? Because the catch data***
***has multiple rows for each night, what is in the weather data for the***
***separate rows on the same night?***

Finally, we'll want to convert our data to a simple count per night using 
`group_by()` and `summarize()`.

```{r}
weather_catch_jan17 <- catches_jan17_full |>
  group_by(collectDate) |>
  summarize(sample_size = n(), # counts number of rows (for each date)
            low_temp = first(lowTemp), # picks the first temp value 
            precip = first(precip)) # picks first precip value

```

Take a moment to view this much smaller data frame. We now have everything
we need to look at sample size versus temperature (or versus precipitation).

```{r}

ggplot(weather_catch_jan17,
       mapping = aes(x = low_temp, y = sample_size)) +
  geom_point() +
  theme_classic()

```

Okay, we've managed some mark-recapture, joining, summarizing and plotting.
Now you can try out a second estimate on your own.

## Western harvest mouse population size (on your own, 15 minutes)

### Your data

This code will help you load the February trapping data, and filter to
just include the western harvest mouse, *Reithrodontomys megalotis*.

```{r}
sjer_feb17 <- read.csv("https://raw.githubusercontent.com/refurrow/WFC70_old/refs/heads/main/data/mam_pertrapnight_2017_02_basic.csv")

rema_feb17 <- filter(sjer_feb17, scientificName== "Reithrodontomys megalotis") |>
  select(collectDate, tagID, recapture, fate, plotID, nlcdClass)
```

### Getting your mark-recapture table

Follow the steps from our *P. boylii* estimate to create a table with
recapture / new capture counts for each night (collectDate). (Hint: `table()` 
function.)

```{r}

# enter your code and comments here

```


Enter your data below, and uncomment and run the lines afterwards. You have
four nights (three resamples), so you'll have three values each for
M, T, and R.

```{r}
#uncomment the lines below and fill in the right values for the
#M, T, and R vectors

#rema_M <- c()
#rema_T <- c()
#rema_R <- c()

#sum(rema_M*rema_T)/sum(rema_R)


```


* ***What is your final estimate for the Feb 2017 population size?***


### More joining (time permitting)

* Create a new data frame that includes all catches for Feb 2017,
* Join it with the weather data, `sjer_weather`,
* Create an identical summary of the data for February (the only 
difference in code will be the name of the initial data frame
you are using),
* use `bind_rows()` to combine the summary from Jan and Feb,
* create a plot of temperature versus sample size for both month's data.

```{r}

# enter your code and comments here

```



## Wrapping up (5 minutes)

Knit your file to HTML, and save the original Rmd, then add them to your WFC 70
folder on Google Drive.


## Extra material

Work on HW 5.

Continue working on the Data Story.

