---
title: 'Lab 7: file management, troubleshooting, data story'
author: "WFC 70, Fall 2025"
output: html_document
---

## Goals

In this lab, we will learn the very basics of file management and 
how to load data into Rmd files when it is stored locally.

By the end of this lab, you should be able to:

-   use "Import Dataset" to load a locally stored file and then copy
the R code into an Rmd code chunk,
-   force a column to be character or numeric data to deal with issues with
how data are loaded,
-   continue exploring your data for the Data Story.

### Resources

Please download from Canvas the file "Lab7_data.zip", and save it to a folder 
on the computer. 

## Part 0: Figure of the week (5 minutes, instructor leading)

--------------------------------------------------------------------------------

--------------------------------------------------------------------------------

## Part 1: Navigating file structure (10 minutes, on your own)

### Unzipping "Lab7_data.zip"

A .zip file is a compressed file that can hold many individual files. It
can be a useful way to share a set of folders with many organized files. 

To decompress it and turn it back into the original folders, you need to
"unzip" it. On a mac, just double-clicking on the folder should automatically 
unzip it to create a folder in the same directory where the .zip file is 
located. On a pc, right click and select "Extract All", then click the
"Extract" button.

At this point, you should now have a folder open with some very poorly named
subfolders (a, b, c, and d). Inside each of these are more folders with
names 1, 2, 3, and 4. And some of those folders have even more folders inside
(i, ii, iii, iv). 

The data are hidden as a csv file within folder d, then subfolder 3, then
subfolder i. Inside of that folder you should find "dispersal_fake.csv". 

### Loading the data

In the upper left of RStudio, click on the File tab, then on "Import Dataset", 
and then "From Text (base)...". This will open a window in which you need
to navigate to the file you want to load. You will need to navigate to 
where you saved the Lab8_data folder, to folder d, then folder 3, then
folder i.  

After you've selected the file, click "Import". It will create
a new data frame in your Environment. ***But more importantly, if you look***
***in the Console in the lower left, you will see the exact R code used to***
***load the data.***

There you will see that you actually just ran a `read.csv()` command, with
a very long string (the "path") with directions to the file you wanted to open.
The backslash at the end of each word or phrase indicates a folder. The start 
of the path will be different for each computer, but the end of the path should
look like "/Lab7_data/d/3/i/dispersal_fake.csv". Those are the directions to
navigate through the Lab8_data folder to get to the actual csv file.

Copy the full code from your Console into this code chunk. You need the data
to be loaded with commands inside of the Rmd if you want to be able to
use that data for more R code later in an Rmd document.

```{r}

library(tidyverse)

# copy and paste the read.csv command that makes your data frame here



# if you are struggling to make that code work, below is code
# to load the data frame from online, so that you can follow
# along with the next section. Uncomment and run it if needed.

#dispersal_fake <- read.csv("https://raw.githubusercontent.com/refurrow/WFC70_old/refs/heads/main/data/dispersal_fake.csv")
```

When you are working with new data for the Data Story, you should create
a folder on your computer where you save the data, and you will need to make
sure you have code within your Rmd that loads the data. Otherwise you will
encounter errors when you try to knit, because the data will not be loaded. 


## Part 2: Dealing with wonky data (10 minutes, on your own)

Let's glimpse our data.

```{r}
# uncomment the line below and run it
# glimpse(dispersal_fake)
```

When you load data into R, there are a few common issues that can arise
with how R tries to automatically identify what kind of data are in each row.

#### Issue 1: text in a numerical column. 

***Always double check that your columns are the data type that you expect.***
Reminder, "num" is numeric, "int" is integer (whole numbers), and "chr" is
character/text (string) data. Sometimes you will have a column that
is mostly numbers, but there are a few entries with text. In our case, the 
spreadsheet used the term "n/a" for missing data rather than "NA". By default,
R assumes that is text. But we can't have a column with multiple data types,
so it will default to the most flexible type. Text/character data is the 
most flexible because it can still store numbers as strings. But you won't 
be able to manipulate them.  You can fix it by forcing a column to be
numeric using `as.numeric()`.

```{r}
# using mutate to fix our distance column
# uncomment and run the lines below

#dispersal_fake <- mutate(dispersal_fake, 
#                         distance = as.numeric(distance))

```

When we used `as.numeric()`, it turned all the strings with numbers
back into numbers, and any other text became NA, which is ideal for our purpose.
So the ominous "Warning" above is just explaining that it did what we wanted --
it introduced NAs where there were non-numbers in the column.

#### Issue 2: missing data.

Sometimes a column in the spreadsheet may have some blank values. By default
for numeric data, these become NAs. You'll notice an example in the 
`age` column, where missing data is now NA. But the blank entry in `brood` is 
just the empty string, "".  So something to watch out for in R is that missing 
data becomes NAs when the column is numeric, but not in columns that are 
character/text data. See Issue 4 for a coding way to deal with that.


#### Issue 3: numbers in a character column.

You should be suspicious of a number in a column that is mostly 
character data -- it might reflect user error. I often
use `table()` on my character variables to double-check what all my 
possible entries are.

```{r}
# uncommon the code below to make the table 

#table(dispersal_fake$brood)
```
Okay, so 4.6 seems like a transcription error or something. Let's look
at the raw data by clicking on its name in the Environment. Hmm, 4.6 is
also the value for the `distance` column for that observation.  Perhaps
this should just be NA, unless we can find the correct `brood` value by 
reviewing our raw data. For making small edits to small spreadsheets, I often 
make a copy of the raw data (called "dispersal_processed.csv", for example), 
and then directly edit that file. If I do that, I add a text file in the same
folder that lists all the editing steps I performed by hand, so I can remember
exactly what I did. Often I will just replace all missing data with NA
by hand as well.

If you do edit your data, make sure you are loading the
newer, processed csv file when you write your code for the data story. Other 
wonky things can happen if R thinks certain columns are in Date format -- it 
can help to force dates to be character data using `as.character()` in some
instances.



#### Issue 4: factors are easier to plot than character data

Categorical data in R will usually be loaded as character data, even when
there are only a few possible levels/values (e.g. "small", "medium" and
"large" for our `brood` column). But there is another data type in R called
factors that is designed to handle categorical data. Converting the relevant
columns to factors can make some aspects of plotting simpler. We won't 
go in depth now, but there is a nice tutorial [linked here: https://r4ds.hadley.nz/factors](https://r4ds.hadley.nz/factors).

```{r}
# example code to remake the brood column with the levels
# small, medium, and large, in that order

# uncomment the code below to try it out

#dispersal_tidy <- dispersal_fake %>% 
#  mutate(brood = factor(brood, levels = c("small", "medium","large")))

```

Basically, factors let you more easily specify which values are listed first,
for controlling things like x-axis order or legend order, and make it easier
to convert missing data to NAs. In the example above, now any values that
were not "small", "medium", or "large" will become NA, and if you made a 
plot with brood as the x-axis aesthetic, it would order the x-axis in the
order your specified (here small, then medium, then large), rather
than alphabetically.


## Data story work (any remaining time)

For the rest of lab, continue your investigation of your data.

1. Make sure you really understand the columns and exactly what they are 
telling you.
2. Get to know the data. Try making a histogram for a column of interest to 
get a sense of the variability. If it's just categorical, use `table()` to
make a table with counts of each value.  Make scatterplots to explore the 
relationship between two numerical variables, and use violin plots to see
how the distribution of a numerical column changes based on the values in
a different categorical column. 
3. Let your curiosity guide you. How can the data tell you interesting things?
Would some background reading help you come up with new questions or 
hypotheses? Now is the time to brainstorm lots of interesting questions
that are answerable with the data.


As you load the data and explore, you may want to code directly in the Data Story 
template Rmd. That way you have your work already in the appropriate place.



## Wrapping up (5 minutes)

Knit any files to HTML, and save the original Rmd (and your Data story template
Rmd, if you have started working in it). Then add these files to your WFC 70
folder on Google Drive.

